---
title: "02_trophoblast"
author: "Victor Yuan"
date: "May 3, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

Here I subset the merged data to only trophoblast cells and perform CCA. 


# 1.0 Libraries 

```{r, message = F, warning = F}
library(tidyr)
library(plyr) 
library(dplyr)
library(Seurat) # V3
library(ggplot2)
library(cowplot)

scobj_merge <- readRDS('../../data/interim/01_scobj_merge.rds')
```

# 2.0 Trophoblast cells

## 2.1 Subset to trophoblast cells
From article:

"The trophoblast clusters (clusters 1, 9, 20, 13 and 16 from Fig. 1d) were taken from the initial analysis of all cells and merged with the enriched EPCAM+ and HLA-G+ cells. The droplet-based and Smart-seq2 datasets were integrated and clustered using the same workflow as described above. Only cells that were identified as trophoblast were considered for trajectory analysis."

So clusters 1, 9, 20, 13, and 16. And also EPCAM+ and HLA-G+ sortd cells.

Note that the legend in figure 1 is incorrect. cluster 13 should be EVT p and cluster 16 should be
EVT, based off MKI67 expression (supplementary figure)

```{r}
Idents(scobj_merge) <- 'final_cluster'
table(Idents(scobj_merge))

troph <- subset(scobj_merge, subset = final_cluster %in% c(1, 9, 20, 13, 16))
troph
table(Idents(troph))
table(troph$platform)

# assign cell type labels to clusters based on paper
troph$celltype <- ifelse(troph$final_cluster == 1, 'VCT',
                         ifelse(troph$final_cluster == 9, 'VCT p',
                                ifelse(troph$final_cluster == 16, 'EVT',
                                       ifelse(troph$final_cluster == 13, 'EVT p',
                                              ifelse(troph$final_cluster == 20, 'SCT', NA)))))

# recreate seurat object, otherwise I receive errors
scobj <- CreateSeuratObject(counts = troph@assays$RNA@counts, meta.data = troph@meta.data)
Idents(scobj) <- 'celltype'
```

## 2.2 CCA

```{r}
scobj_list <- SplitObject(object = scobj, split.by = "platform")

# find top 2000 variable features
for (i in 1:length(scobj_list)) {
  scobj_list[[i]] <- NormalizeData(object = scobj_list[[i]], verbose = T)
  scobj_list[[i]] <- FindVariableFeatures(object = scobj_list[[i]],
                                          selection.method = "vst", 
                                          nfeatures = 2000, 
                                          verbose = T)
}

# Merge datasets
reference_list <- scobj_list[c("10x", "DropSeq")]
anchors <- FindIntegrationAnchors(object.list = reference_list, dims = 1:30) # find anchors
troph_merge <- IntegrateData(anchorset = anchors, dims = 1:30,
                             features.to.integrate = rownames(scobj_list[[1]])) #
```

## 2.3 PCA, UMAP

```{r}
#scobj_merge <- ScaleData(object = scobj_merge, verbose = FALSE)
troph_merge <- ScaleData(object = troph_merge, do.scale = F, do.center = T, verbose = T)
troph_merge <- RunPCA(object = troph_merge, npcs = 30, verbose = FALSE)
p1 <- DimPlot(object = troph_merge, reduction = "pca", group.by = "celltype")
p2 <- DimPlot(object = troph_merge, reduction = "pca", group.by = "celltype", dims = c(3,4))
p3 <- DimPlot(object = troph_merge, reduction = "pca", group.by = "celltype", dims = c(5,6))
p4 <- DimPlot(object = troph_merge, reduction = "pca", group.by = "celltype", dims = c(7,8))
plot_grid(p1,p2,p3,p4)

troph_merge <- RunUMAP(object = troph_merge, reduction = "pca", dims = 1:5)

p1 <- DimPlot(object = troph_merge, reduction = "umap", group.by = "platform")
p2 <- DimPlot(object = troph_merge, reduction = "umap", group.by = "celltype", 
    label = TRUE, repel = TRUE) + NoLegend()
plot_grid(p1, p2)
```

saveRDS(troph_merge, '../../data/interim/02_troph_merge.rds')