---
title: "10_analysis"
author: "Victor Yuan"
date: "April 21, 2019"
output:
  html_document:
    keep_md: true
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: spacelab
editor_options: 
  chunk_output_type: console
---

This is for analysing the scRNAseq data
# 1.0 Libraries and Data

```{r}
library(readxl) # read data in
library(plyr) 
library(stringr)
library(knitr);library(kableExtra) # for displaying pretty tables
library(dplyr)
library(Seurat) # V3
library(readr)
library(ggplot2)
library(cowplot)
library(tidyr)

scobj_merge <- readRDS('../../data/interim/01_scobj_merge.rds')
```

# 2.0 Validate hits

First subset to trophoblast

## 2.1 Subset to Trophoblast Cells

From article:

"The trophoblast clusters (clusters 1, 9, 20, 13 and 16 from Fig. 1d) were taken from the initial analysis of all cells and merged with the enriched EPCAM+ and HLA-G+ cells. The droplet-based and Smart-seq2 datasets were integrated and clustered using the same workflow as described above. Only cells that were identified as trophoblast were considered for trajectory analysis."

So clusters 1, 9, 20, 13, and 16. And also EPCAM+ and HLA-G+ sortd cells.

```{r}
Idents(scobj_merge) <- 'final_cluster'
table(Idents(scobj_merge))

troph <- subset(scobj_merge, subset = final_cluster %in% c(1, 9, 20, 13, 16))
troph
table(Idents(troph))
table(troph$platform)

# assign cell type labels to clusters based on paper
troph$celltype <- ifelse(troph$final_cluster == 1, 'VCT',
                         ifelse(troph$final_cluster == 9, 'VCT p',
                                ifelse(troph$final_cluster == 13, 'EVT',
                                       ifelse(troph$final_cluster == 16, 'EVT p',
                                              ifelse(troph$final_cluster == 20, 'SCT', NA)))))
Idents(troph) <- 'celltype'
```

Recompute pca and umap coordinates on trophoblast cells:

```{r}
#redo umap / pca
troph <- ScaleData(object = troph, verbose = FALSE)
troph <- RunPCA(object = troph, npcs = 10, verbose = FALSE)
troph <- RunUMAP(object = troph, reduction = "pca", 
    dims = 1:10)
DimPlot(object = troph, reduction = "umap", group.by = "celltype") + 
  theme_classic() +
  ggtitle('Trophoblast cells (n = 14366)')
```

then plot expression of hits

## 2.2 list of hits

### Violin plots

```{r}
hitlist <- read_xlsx('../../2019-03-19 currated scRNA-seq gene list.xlsx')
hitlist <- hitlist %>% gather(key = 'Geneset', value = 'Gene') %>% filter(!is.na(Gene))

# order levels to plot by
Idents(troph) <- factor(as.character(Idents(troph)), 
                        levels = c('VCT', 'VCT p', 'EVT', 'EVT p', 'SCT'))

#violin plots
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Troph (general)'], 
        pt.size = F, ncol = 4, same.y.lims =  T) 
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='CTB'],
        pt.size = F, same.y.lims =  T)
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Dist CCTB'], 
        pt.size = F, same.y.lims =  T)
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Prox CCTB'],
        pt.size = F, same.y.lims =  T)
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Syn Pre CTB'],
        pt.size = F, same.y.lims =  T)
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Target High'],
        pt.size = F, same.y.lims =  T, ncol = 5)
VlnPlot(object = troph, features = hitlist$Gene[hitlist$Geneset=='Target Low'],
        pt.size = F, same.y.lims =  T, ncol = 5)
```

### heatmap

```{r}
#plot average expression
average_exp <- AverageExpression(object = troph, return.seurat = T)
ave_exp_mat <- average_exp@assays$integrated@scale.data[,] %>% as.data.frame()

library(pheatmap)
#create annotation dataframe for genes
row_label <- as.data.frame(hitlist) 
row_label$Gene <- make.unique(row_label$Gene)
rownames(row_label) <- gsub('\\.', '-', row_label$Gene)
row_label$Geneset <- factor(row_label$Geneset, 
                            levels = c('Troph (general)', 'CTB', 'Dist CCTB', 'Prox CCTB', 
                                       'Syn Pre CTB', 'Target High', 'Target Low'))

row_label <- row_label[intersect(rownames(row_label), rownames(ave_exp_mat)),]
ave_exp_mat <- ave_exp_mat[rownames(row_label),]
pheatmap(ave_exp_mat[,c('VCT', 'VCT p', 'EVT', 'EVT p', 'SCT')], 
         cluster_rows = F, cluster_cols = F, 
         annotation_row = row_label[,'Geneset',F],
         color = RColorBrewer::brewer.pal(9, "Blues"),
         cellwidth = 15, cellheight = 15)

```